---
title: "First look "
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
  - \usepackage{bm}
  - \usepackage{xcolor}
  - \usepackage{amssymb}
output: 
  html_document:
    df_print: paged
    theme: flatly
    highlight: tango #pygments
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
    smooth_scroll: true
    code_fold: hide
  urlcolor: blue
---
  

```{r setup, include=FALSE, message=F, echo=F, warning=F}
# LIBRARIES----
library(tidyverse)
library(gridExtra)
library(here)
library(knitr)


# DOCUMENT SETUP ----
# detect pdf/html output, set chunk options, sci.notation 
latex_out <- knitr::is_latex_output()
knitr::opts_chunk$set(cache=F, 
                      message=F, 
                      echo=!latex_out, 
                      warning=F, 
                      fig.height=4, 
                      fig.width=6)


# PLOTTING FCNS----
# default settings; transparency for base R
pardefault <- par(no.readonly=T)
palramp <- colorRampPalette(c("deepskyblue", "purple", "orangered"))
transp_pal <- function(pal, alpha=75){
  # add transparency to palette 'pal'
  return( rgb(t(col2rgb(pal)), alpha = alpha, maxColorValue = 255) )
} 

#ggplot2 colors into base R
gg_color <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}



# SIGNAL/PROCESSING ----
# Apply affine transformation to y so that max(y) = 1 and min(y) = -1
# We'll need this later to make sure input to play() is valid
# Anything above +1/below -1 will be clipped, resulting in distortion
wave_norm <- function(y) {
  bias = mean(y)
  y <- y - bias
  M <- max(y)
  m <- min(y)
  scale <- if (M > abs(m)) M else abs(m)
  y <- y / scale
  return(y)
}




# TEXT/TABLE FORMATTING----
colorize <- function(x, color="red") {
  # text color conditional on latex/html output
  # from rmarkdown cookbook
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{ %s}{ %s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, x)
  } else x
}

# kable NA handling
options(knitr.kable.NA = '')

# mykable function
mykable <- function(tab, cap,
                    latex_options=c("hold_position", "scale_down", "striped"), 
                    bootstrap_options=c("striped", "hover", "condensed"), 
                    full_width=F, position="center", ...){
  # kable formatting conditional on latex or html output
  if (knitr::is_latex_output()){
    kable(x=tab, caption=cap, ...) %>%
      kableExtra::kable_styling(latex_options = latex_options)
  } else if (knitr::is_html_output()){
    kable(x=tab, caption=cap, ...) %>%
      kableExtra::kable_styling(bootstrap_options = bootstrap_options, full_width=full_width, position=position)
  }
}

# SOURCE common functions ----
# make a %notin% function to mirror %in%
`%notin%` <- Negate(`%in%`)
```










# `audio` package

## fooling around

play a pitch, same pitch plus another

```{r foolin}
library(audio)

pal <- transp_pal(palramp(4), alpha = 175)
palette(pal)

# continuous pitch
s1 <- sin(1:88000/10)
plot(s1[1:500], pch=',', main = "wave `s1`", col=1)
play(s1)

# another continuous pitch
s2 <- sin(1:88000/20)
plot(s2[1:500], pch=',', main = "wave `s2`", col=1)
play(s2)


# play together
plot(c(s1+s2)[1:500], pch=',', main = "wave s1 + s2", col=1)
play(wave_norm(s1 + s2))

# 2 sec white noise
wn <- rnorm(88000, 0, 1)
plot(wn[1:500], pch=',', col=1)
play(wave_norm(wn))

# adding sound wave to white noise --> white noise
play(wave_norm(s1 + wn))

```





## middle A (A4)

1 second = 44k length vector

middle A: 44o Hz.  i.e. need to cycle through 2pi every 100 elements?

```{r middleA}
# 1 second of A4
mid_a <- rep(seq(0, 2*pi,length.out=100), 440)
play(mid_a)
```



## Piano key frequencies

[Piano key frequencies](https://en.wikipedia.org/wiki/Piano_key_frequencies) gives us a mapping from the key number on a standard piano to hz_:

$$f(n) = 2^{\frac{n-49}{12}} \times 440 \text{ Hz}$$


__Write some functions:__

+ `pkey_to_hz(keynum)` maps piano key number to Hz
+ `hz_to_1sec(hz)` generates 44k-length vector of sine wave with specified frequency (1 second)
+ `note(pkey, vol=1)` combines them, adds volume multiplier (amplitude = volume).  
  + can feed `note()` output directly into `play()` from `audio` package. 


```{r basic_funcs}
pkey_to_hz <- function(keynum){
  # maps piano key to hz
  2^((keynum-49) / 12) * 440
}

hz_to_1sec <- function(hz){
  # generates 440k length sine wave vector with frequency = hz
  rep(seq(0, 2*pi, length.out = floor(44000/hz)), floor(hz))
}

note <- function(pkey, vol=1){
  # combines basic funcs to output playable vectors
  vol*hz_to_1sec(hz = pkey_to_hz(keynum = pkey))
}
```



__Test:__

```{r play_basic}
play(
  hz_to_1sec(pkey_to_hz(49))
)
play(note(49, 1))
play(note(49, .25))
play(note(49, .1))


```

__volume specification doesn't really work?__  vol=0.25 seems louder than vol=1 ... 





## play an A major scale

_Note;_ Needed to revamp function generating 44k length vector of specified hz because

+ rep() doesn't vectorize well
+ rep() needs integer-valued arguments


```{r}

hz_to_1sec <- function(hz){
  # generates 440k length sine wave vector with frequency = hz
  outlength <- floor(44000/hz)
  reps <- floor(hz)

  unlist(mapply(out=outlength, 
         reps=reps, 
         function(out, reps) 
           rep(seq(0, 2*pi, length.out = out), reps)
         ))
 
  # floor arguments will make these slightly sharp
  # and slightly shorter than 1 second
}


Ascale <- c(49, 51, 53, 54, 56, 58, 60, 61)
play(note(Ascale))

```

__Hey, it works!!__



But volume isn't working =(:

```{r}
# testing out volume now
play(note(
  Ascale, 
  vol=seq(.1:1, length.out = length(Ascale))
  )
)
```


Trying out chords
```{r}
a4vec <- note(Ascale[1:6])
vec2 <- note(Ascale[3:8])

# this is kind of nice when executed together
# (select both and hit ctrl + Enter)
play(a4vec)
play(vec2)


# yikes!  no bueno!  be warned, it's terrible.
play(a4vec+vec2)

```



# `tuneR` package













